---
title : OzoneApp
subtitle    : Ozone pollution prediction and global warming
author : Géraud Dugé de Bernonville
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
```


## Intro

* Global warming is a major preoccupation, and recent studies show that temperature raising is
a real concern.
* Ozone pollution is another major concern as our populations became more and more urban, and we can see that Ozone pollution alert are more and more frequent.
* [__OzoneApp__](http://geraudster.shinyapps.io/ozoneApp) aims to analyze the impact of global warming on Ozone pollution and will help predicting Ozone pollution depending on temperature raise.

We use the _airquality_ dataset provided by R, this dataset contains Daily air quality measurements in New York, May to September 1973.

--- .class #id 

## Dataset

The dataset consists of 154 observations on 6 variables:

* `Ozone`: Mean ozone in parts per billion from 1300 to 1500 hours at Roosevelt Island
* `Solar.R`: Solar radiation in Langleys in the frequency band 4000–7700 Angstroms from 0800 to 1200 hours at Central Park
* `Wind`: Average wind speed in miles per hour at 0700 and 1000 hours at LaGuardia Airport
* `Temp`: Maximum daily temperature in degrees Fahrenheit at La Guardia Airport.

Below is a sample code after imputing missing values:
```{r, echo=FALSE}
data(airquality)

library(mice)
# Impute missing values
imputed <- mice(airquality, m = 1, print = FALSE)
airquality$Solar.R[is.na(airquality$Solar.R)] <- imputed$imp$Solar.R[[1]]
airquality$Ozone[is.na(airquality$Ozone)] <- imputed$imp$Ozone[[1]]
head(airquality)
airquality2 <- airquality
```

--- .class #id 

## Data exploration

* Ozone distribution is left skewed so we will use the log of Ozone to have a normal distribution.

<!--
```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 3}
library(reshape2)
airquality2$LogOzone <- log(airquality2$Ozone)
ozonemelt <- melt(airquality2[,c('Ozone', 'LogOzone')])
qplot(value, data = ozonemelt, geom = 'histogram') + facet_grid(facets = . ~ variable,  scales = 'free_x')
```
-->

* The following correlation matrix shows that Wind and Temperature have strong correlation with Ozone:

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=4}
library(reshape2)
airquality.cor <- cor(airquality2[, c('LogOzone', 'Solar.R', 'Temp', 'Wind')])
airquality.cor[lower.tri(airquality.cor)] <- NA
airquality.cor.melt <- melt(airquality.cor)
levels(airquality.cor.melt$Var1) <- list(LogOzone = 'LogOzone', Solar.R = 'Solar.R', Temp = 'Temp', Wind = 'Wind')
levels(airquality.cor.melt$Var2) <- rev(list(LogOzone = 'LogOzone', Solar.R = 'Solar.R', Temp = 'Temp', Wind = 'Wind'))

qplot(x=Var2, y=Var1, data=airquality.cor.melt, fill=value, geom="tile") +
  geom_text(aes(Var2, Var1, label = round(value,2))) +
  scale_fill_gradient(name=expression('Correlation'),
                      limits = c(-1, 1),
                      low = "steelblue", high = "orange", na.value = 'white') +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), 
        panel.grid.major = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      axis.ticks = element_blank(),
      legend.justification = c(1, 0),
      legend.position = c(0.9, 0.7),
      legend.direction = "horizontal")

```


--- .class #id 

## The model for prediction

Below we compare different linear regression models:
```{r, fig.height=3}
library(caret)
models <- list(Temp = train(LogOzone ~ Temp, airquality2, method = 'lm'),
               'Temp + Wind' = train(LogOzone ~ Temp + Wind, airquality2, method = 'lm'),
               'Temp + Wind + Solar.R' = train(LogOzone ~ Temp + Wind + Solar.R, airquality2, method = 'lm'))
bwplot(resamples(models))

```

We chose the third (`Temp + Wind + Solar.R`), which has lower RMSE and greater R^2.

--- .class #id 

## How to use

Bla bla 

"Ozongassmolekyl" by Ingvald Straume - Own work. Licensed under CC BY-SA 3.0 via Wikimedia Commons - http://commons.wikimedia.org/wiki/File:Ozongassmolekyl.png#mediaviewer/File:Ozongassmolekyl.png
