# Analysis of relationship between car transmission and fuel consumption

## Synopsis

We will use the mtcars dataset to determine the kind of transmission (automatic/manual)
that has the most Miles per Gallon, and in which proportion.

## Data processing

We will use the mtcars dataset included in the R core.

Format is: 
    
* TODO

First we load the data:

```{r}
data(mtcars)
names(mtcars)
nrow(mtcars)
head(mtcars)
```

Next we prepare the data (to have proper label when drawing plots):
```{r}
mtcars <- transform(mtcars, transmission = factor(mtcars$am,
                                                     levels = c(0,1),
                                                     labels = c("Automatic", "Manual")))
```

## Exploratory analysis

We would like to determine if there is a relationship between transmission an mpg.

```{r}
boxplot(mpg~transmission,data=mtcars, main="Car Milage Data",
   xlab="Transmission type", ylab="Miles Per Gallon") 
```

So it seems there is a relation between transmission and mpg: cars use more fuel when
transmission is manual than automatic

```{r}
cor(mtcars$am, mtcars$mpg)

```

## Regression model

```{r}
model <- lm(mpg~am, data=mtcars)
model.summary <- summary(model)
model.summary
```

This model shows that:

* the coefficient is statistically significant (with a p-value of `r model.summary$coefficients[2,4]`)
* transmission explains `r model.summary$r.squared * 100`% of the variance
* when the transmision change from automatic (0) to manual(1), there is an increase in mpg of
`r model$coefficients['am']`

Next the confidence interval:

```{r}
confint(model)
```

Let's plot the residuals:

```{r}
plot(resid(model))
```